---
title: "Benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sceua)

benchmark_packages_available <-
  requireNamespace("rtop", quietly = TRUE) &&
  requireNamespace("SoilHyP", quietly = TRUE) &&
  requireNamespace("bench", quietly = TRUE)
```

The `sceua` package is not the first R package with the Shuffled Complex
Evolution algorithm for global optimisation. To our knowledge, the
[`rtop`](https://cran.r-project.org/package=rtop) and
[`SoilHyP`](https://cran.r-project.org/web/packages/SoilHyP/index.html)
packages have implemented this algorithm before. However, `sceua` provides the
fastest implementation available in R. See benchmarks below.

## A non-linear curve-fitting problem

To test the three packages against each other, we first use the same non-linear
curve-fitting problem as in the `rtop::sceua()` documentation. We generate
noisy observations from a sine function with three parameters and recover them
by minimising the sum of squared errors.

```{r sine-data}
set.seed(1)

fun <- function(x, pars) pars[2] * sin(x * pars[1]) + pars[3]

x <- rnorm(50, sd = 3)
y <- fun(x, pars = c(5, 2, 3)) + rnorm(length(x), sd = 0.3)

OFUN <- function(pars, x, yobs) {
  yvals <- fun(x, pars)
  sum((yvals - yobs)^2)
}
```

## Timing comparison

The `rtop` and `SoilHyP` packages contain SCE-UA implementations in pure R.
Below is a comparison of these two packages with the Rust-backed `sceua`
package on the same problem.

> `rtop::sceua()` and `SoilHyP::SCEoptim()` print convergence information to
> the console. The helper below captures that output and returns only the
> optimised value.

```{r benchmark-setup, eval = benchmark_packages_available}
#| code-fold: true
silent_rtop <- function(...) {
  utils::capture.output(
    res <- rtop::sceua(...),
    type = "output"
  )
  res$value
}

silent_soilhyp <- function(...) {
  utils::capture.output(
    res <- SoilHyP::SCEoptim(...),
    type = "output"
  )
  res$value
}

# The first `rtop` call in a fresh session can land in a different local basin;
# a single warm-up call makes the benchmarked calls comparable to `sceua`.
invisible(silent_rtop(
  OFUN,
  pars = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  maxn = 10000
))

invisible(silent_soilhyp(
  FUN = OFUN,
  par = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  control = list(maxeval = 10000)
))
```

```{r benchmark, eval = benchmark_packages_available}
set.seed(1)

bench::mark(
  sceua = sceua::sceua(
    fn = OFUN,
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    initial = c(0.1, 0.1, 0.1),
    x = x,
    yobs = y,
    max_evaluations = 10000
  )$value,
  rtop = silent_rtop(
    OFUN,
    pars = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    maxn = 10000
  ),
  soilhyp = silent_soilhyp(
    FUN = OFUN,
    par = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    control = list(maxeval = 10000)
  ),
  iterations = 25,
  relative = FALSE,
  check = FALSE
)
```

## Q. Duan's test-function benchmarks

The following benchmarks use R translations of the Q. Duan test functions
published alongside the [Matlab code](https://www.mathworks.com/matlabcentral/fileexchange/7671-shuffled-complex-evolution-sce-ua-method).
We use one aligned strict configuration for all problems: objective
convergence is disabled (`pcento = 0`) and parameter convergence is tightened
to `1e-6`. The same seed is used for every package so the initial populations
are comparable.

Each problem below gets its own table with the best value found, the known
optimum, the mean absolute error (`MAE` = `|value - optimum|`), the median
wall-clock time over 5 timed runs, and a `Timing` bar scaled to the slowest
run within that problem (a full bar marks the slowest package). Every package
is timed on every problem; cells show `—` only when a package errors or fails
to return a finite value. Per-problem bounds are visible in the folded code
chunk below.

```{r duan-test-functions, eval = benchmark_packages_available}
#| code-fold: true
goldstein_price <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  u1 <- (x1 + x2 + 1)^2
  u2 <- 19 - 14 * x1 + 3 * x1^2 - 14 * x2 + 6 * x1 * x2 + 3 * x2^2
  u3 <- (2 * x1 - 3 * x2)^2
  u4 <- 18 - 32 * x1 + 12 * x1^2 + 48 * x2 - 36 * x1 * x2 +
    27 * x2^2
  (1 + u1 * u2) * (30 + u3 * u4)
}

rosenbrock <- function(x) {
  100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
}

six_hump_camelback <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  (4 - 2.1 * x1^2 + x1^4 / 3) * x1^2 + x1 * x2 +
    (-4 + 4 * x2^2) * x2^2
}

rastrigin_duan <- function(x) {
  x[1]^2 + x[2]^2 - cos(18 * x[1]) - cos(18 * x[2])
}

griewank_duan <- function(x) {
  divisor <- if (length(x) == 2L) 200 else 4000
  sum(x^2 / divisor) - prod(cos(x / sqrt(seq_along(x)))) + 1
}

shekel <- function(x) {
  a <- rbind(
    c(4, 1, 8, 6, 3, 2, 5, 8, 6, 7),
    c(4, 1, 8, 6, 7, 9, 5, 1, 2, 3.6),
    c(4, 1, 8, 6, 3, 2, 3, 8, 6, 7),
    c(4, 1, 8, 6, 7, 9, 3, 1, 2, 3.6)
  )
  c_values <- c(0.1, 0.2, 0.2, 0.4, 0.4, 0.6, 0.3, 0.7, 0.5, 0.5)

  -sum(vapply(seq_along(c_values), function(i) {
    1 / (sum((x - a[, i])^2) + c_values[i])
  }, numeric(1)))
}

hartman <- function(x) {
  a6 <- rbind(
    c(10, 0.05, 3, 17),
    c(3, 10, 3.5, 8),
    c(17, 17, 1.7, 0.05),
    c(3.5, 0.1, 10, 10),
    c(1.7, 8, 17, 0.1),
    c(8, 14, 8, 14)
  )
  c6 <- c(1, 1.2, 3, 3.2)
  p6 <- rbind(
    c(0.1312, 0.2329, 0.2348, 0.4047),
    c(0.1696, 0.4135, 0.1451, 0.8828),
    c(0.5569, 0.8307, 0.3522, 0.8732),
    c(0.0124, 0.3736, 0.2883, 0.5743),
    c(0.8283, 0.1004, 0.3047, 0.1091),
    c(0.5886, 0.9991, 0.6650, 0.0381)
  )

  -sum(vapply(seq_along(c6), function(i) {
    c6[i] * exp(-sum(a6[, i] * (x - p6[, i])^2))
  }, numeric(1)))
}

duan_control <- list(
  seed = 1L,
  complexes = 5L,
  max_evaluations = 10000L,
  kstop = 5L,
  pcento = 0,
  parameter_epsilon = 1e-6
)

duan_problems <- list(
  goldstein_price = list(
    fn = goldstein_price,
    lower = c(-2, -2),
    upper = c(2, 2),
    initial = c(1, 1),
    optimum = 3
  ),
  rosenbrock = list(
    fn = rosenbrock,
    lower = c(-5, -5),
    upper = c(5, 5),
    initial = c(-3, 4),
    optimum = 0
  ),
  six_hump_camelback = list(
    fn = six_hump_camelback,
    lower = c(-5, -2),
    upper = c(5, 8),
    initial = c(0, 0),
    optimum = -1.031628453489877
  ),
  rastrigin_duan = list(
    fn = rastrigin_duan,
    lower = c(-1, -1),
    upper = c(1, 1),
    initial = c(0.5, 0.5),
    optimum = -2
  ),
  griewank_duan = list(
    fn = griewank_duan,
    lower = rep(-600, 10),
    upper = rep(600, 10),
    initial = rep(100, 10),
    optimum = 0
  ),
  shekel = list(
    fn = shekel,
    lower = rep(0, 4),
    upper = rep(10, 4),
    initial = rep(1, 4),
    optimum = -10.5364098252
  ),
  hartman = list(
    fn = hartman,
    lower = rep(0, 6),
    upper = rep(1, 6),
    initial = rep(0.5, 6),
    optimum = -3.322368011415515
  )
)

sceua_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  sceua::sceua(
    fn = problem$fn,
    lower = problem$lower,
    upper = problem$upper,
    initial = problem$initial,
    max_evaluations = duan_control$max_evaluations,
    kstop = duan_control$kstop,
    pcento = duan_control$pcento,
    complexes = duan_control$complexes,
    points_per_complex = 2L * n + 1L,
    simplex_size = n + 1L,
    evolution_steps = 2L * n + 1L,
    min_complexes = duan_control$complexes,
    parameter_epsilon = duan_control$parameter_epsilon
  )$value
}

rtop_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  silent_rtop(
    problem$fn,
    pars = problem$initial,
    lower = problem$lower,
    upper = problem$upper,
    maxn = duan_control$max_evaluations,
    kstop = duan_control$kstop,
    pcento = duan_control$pcento,
    ngs = duan_control$complexes,
    npg = 2L * n + 1L,
    nps = n + 1L,
    nspl = 2L * n + 1L,
    mings = duan_control$complexes,
    peps = duan_control$parameter_epsilon,
    iniflg = 1
  )
}

soilhyp_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  silent_soilhyp(
    FUN = problem$fn,
    par = problem$initial,
    lower = problem$lower,
    upper = problem$upper,
    control = list(
      maxeval = duan_control$max_evaluations,
      ncomplex = duan_control$complexes,
      cce.iter = 2L * n + 1L,
      reltol = duan_control$pcento,
      tolsteps = duan_control$kstop,
      initsample = "latin"
    )
  )
}

safe_duan_value <- function(package, problem) {
  value <- tryCatch(
    switch(
      package,
      sceua = sceua_duan_value(problem),
      rtop = rtop_duan_value(problem),
      soilhyp = soilhyp_duan_value(problem)
    ),
    error = function(e) NA_real_
  )

  as.numeric(value)
}

# Number of timed runs per (problem, package). bench::mark runs each
# expression exactly this many times.
duan_iterations <- 5L

# Dependency-free cell visual: a fixed-width unicode bar for a scalar value
# scaled to a per-problem maximum. A positive sub-maximum value always gets
# at least one filled cell so the fastest run still shows a sliver.
bar_text <- function(value, max_value, width = 10L,
                     fill = "\u2588", empty = "\u2591") {
  if (is.na(value) || is.na(max_value) || max_value == 0) return("")
  n <- round(width * value / max_value)
  if (value > 0 && value < max_value) n <- max(n, 1L)
  n <- min(max(n, 0L), width)
  paste0(strrep(fill, n), strrep(empty, width - n))
}

fmt_time <- function(seconds) {
  if (is.na(seconds)) return("\u2014")
  if (seconds >= 1) return(sprintf("%.2f s", seconds))
  if (seconds >= 1e-3) return(sprintf("%.2f ms", seconds * 1e3))
  sprintf("%.0f \u00b5s", seconds * 1e6)
}

# Compute the best value for every (problem, package) pair. `safe_duan_value`
# returns NA when a package errors, so a failed run still has a row.
packages <- c("sceua", "rtop", "soilhyp")
duan_values <- do.call(rbind, lapply(names(duan_problems), function(name) {
  problem <- duan_problems[[name]]
  values <- vapply(packages, safe_duan_value, numeric(1), problem = problem)
  data.frame(
    problem = name,
    package = packages,
    value = values,
    abs_error = abs(values - problem$optimum),
    stringsAsFactors = FALSE
  )
}))

# Time every pair. `check = FALSE` disables bench's result-equality check; the
# tables below compare values to the optimum explicitly. A pair that errors is
# recorded with `NA` median time rather than aborting the vignette.
benchmark_duan_row <- function(name, package, problem) {
  exprs <- list(
    sceua = quote(sceua_duan_value(problem)),
    rtop = quote(rtop_duan_value(problem)),
    soilhyp = quote(soilhyp_duan_value(problem))
  )
  set.seed(duan_control$seed)
  res <- tryCatch(
    bench::mark(
      eval(exprs[[package]]),
      iterations = duan_iterations,
      relative = FALSE,
      check = FALSE
    ),
    error = function(e) NULL
  )
  median_time <- if (is.null(res)) NA_real_ else as.numeric(res$median[[1]])
  data.frame(
    problem = name,
    package = package,
    median_time = median_time,
    stringsAsFactors = FALSE
  )
}

duan_timing <- do.call(rbind, lapply(names(duan_problems), function(name) {
  problem <- duan_problems[[name]]
  do.call(rbind, lapply(packages, function(pkg) {
    benchmark_duan_row(name, pkg, problem)
  }))
}))

duan_summary <- merge(
  duan_values,
  duan_timing,
  by = c("problem", "package"),
  sort = FALSE
)
duan_summary <- duan_summary[order(
  match(duan_summary$problem, names(duan_problems)),
  match(duan_summary$package, packages)
), ]
duan_summary$optimum <- vapply(
  duan_summary$problem,
  function(n) duan_problems[[n]]$optimum,
  numeric(1)
)

fmt_value <- function(x) ifelse(is.na(x), "\u2014", sprintf("%.6g", x))
fmt_err <- function(x) ifelse(is.na(x), "\u2014", sprintf("%.3e", x))

# Per-problem table. The Timing bar is scaled to the slowest run within the
# problem, so a full bar marks the slowest package and shorter bars mark
# faster ones. A package that errors shows em dashes.
duan_problem_table <- function(name) {
  rows <- duan_summary[duan_summary$problem == name, , drop = FALSE]
  max_time <- if (all(is.na(rows$median_time))) {
    NA_real_
  } else {
    max(rows$median_time, na.rm = TRUE)
  }
  timing_bar <- vapply(
    rows$median_time, bar_text, character(1), max_value = max_time
  )

  data.frame(
    Package = rows$package,
    `Best value` = fmt_value(rows$value),
    Optimum = sprintf("%.6g", rows$optimum[1]),
    MAE = fmt_err(rows$abs_error),
    `Median time` = vapply(rows$median_time, fmt_time, character(1)),
    Timing = ifelse(timing_bar == "", "\u2014", timing_bar),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
}

duan_problem_caption <- function(name) {
  p <- duan_problems[[name]]
  sprintf(
    "%s: bounds [%s], optimum %g, %d parameters",
    name,
    paste(sprintf("[%g, %g]", p$lower, p$upper), collapse = ", "),
    p$optimum,
    length(p$lower)
  )
}
```

```{r duan-summary, results='asis', eval = benchmark_packages_available}
for (name in names(duan_problems)) {
  cat("###", name, "\n\n")
  cat(knitr::kable(
    duan_problem_table(name),
    align = c("l", "r", "r", "r", "r", "l"),
    caption = duan_problem_caption(name)
  ), sep = "\n")
  cat("\n\n")
}
```
