---
title: "Benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sceua)
```

The `sceua` package is not the first R package with the Shuffled Complex
Evolution algorithm for global optimisation. To our knowledge, the
[`rtop`](https://cran.r-project.org/package=rtop) and
[`SoilHyP`](https://cran.r-project.org/web/packages/SoilHyP/index.html)
packages have implemented this algorithm before. However, `sceua` provides the
fastest implementation available in R! See benchmarks below.

## A non-linear curve-fitting problem

To test the three packages against each other, we use the same non-linear
curve-fitting problem as in the `rtop::sceua()` documentation. We generate
noisy observations from a sine function with three parameters and recover them
by minimising the sum of squared errors.

```{r sine-data}
set.seed(1)

fun <- function(x, pars) pars[2] * sin(x * pars[1]) + pars[3]

x <- rnorm(50, sd = 3)
y <- fun(x, pars = c(5, 2, 3)) + rnorm(length(x), sd = 0.3)

OFUN <- function(pars, x, yobs) {
  yvals <- fun(x, pars)
  sum((yvals - yobs)^2)
}
```

## Timing comparison

The `rtop` and `SoilHyP` packages contain SCE-UA implementations in pure R.
Below is a comparison of these two packages with the Rust-backed `sceua`
package on the same problem.

> `rtop::sceua()` and `SoilHyP::SCEoptim()` print convergence information to the console. The helper
> below captures that output and returns only the optimised value.

```{r benchmark-setup, eval = requireNamespace("rtop", quietly = TRUE) && requireNamespace("SoilHyP", quietly = TRUE) && requireNamespace("bench", quietly = TRUE)}
#| code-fold: true
silent_rtop <- function(...) {
  utils::capture.output(
    res <- rtop::sceua(...),
    type = "output"
  )
  res$value
}

silent_soilhyp <- function(...) {
  utils::capture.output(
    res <- SoilHyP::SCEoptim(...),
    type = "output"
  )
  res$value
}

# The first `rtop` call in a fresh session can land in a different local basin;
# a single warm-up call makes the benchmarked calls comparable to `sceua`.
invisible(silent_rtop(
  OFUN,
  pars = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  maxn = 10000
))

invisible(silent_soilhyp(
  FUN = OFUN,
  par = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  control = list(maxeval = 10000)
))
```

```{r benchmark, eval = requireNamespace("rtop", quietly = TRUE) && requireNamespace("SoilHyP", quietly = TRUE) && requireNamespace("bench", quietly = TRUE)}
bench::mark(
  sceua = sceua::sceua(
    fn = OFUN,
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    initial = c(0.1, 0.1, 0.1),
    x = x,
    yobs = y,
    max_evaluations = 10000,
    seed = 1
  )$value,
  rtop = silent_rtop(
    OFUN,
    pars = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    maxn = 10000
  ),
  soilhyp = silent_soilhyp(
    FUN = OFUN,
    par = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    control = list(maxeval = 10000)
  ),
  iterations = 15,
  relative = FALSE,
  check = function(x, y) isTRUE(all.equal(x, y, tolerance = 1e-3))
)
```

Yay! 60-80 times faster than predecessors 🚂→🚅
