---
title: "Benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sceua)

benchmark_packages_available <-
  requireNamespace("rtop", quietly = TRUE) &&
  requireNamespace("SoilHyP", quietly = TRUE) &&
  requireNamespace("bench", quietly = TRUE)
```

The `sceua` package is not the first R package with the Shuffled Complex
Evolution algorithm for global optimisation. To our knowledge, the
[`rtop`](https://cran.r-project.org/package=rtop) and
[`SoilHyP`](https://cran.r-project.org/web/packages/SoilHyP/index.html)
packages have implemented this algorithm before. However, `sceua` provides the
fastest implementation available in R. See benchmarks below.

## A non-linear curve-fitting problem

To test the three packages against each other, we first use the same non-linear
curve-fitting problem as in the `rtop::sceua()` documentation. We generate
noisy observations from a sine function with three parameters and recover them
by minimising the sum of squared errors.

```{r sine-data}
set.seed(1)

fun <- function(x, pars) pars[2] * sin(x * pars[1]) + pars[3]

x <- rnorm(50, sd = 3)
y <- fun(x, pars = c(5, 2, 3)) + rnorm(length(x), sd = 0.3)

OFUN <- function(pars, x, yobs) {
  yvals <- fun(x, pars)
  sum((yvals - yobs)^2)
}
```

## Timing comparison

The `rtop` and `SoilHyP` packages contain SCE-UA implementations in pure R.
Below is a comparison of these two packages with the Rust-backed `sceua`
package on the same problem.

> `rtop::sceua()` and `SoilHyP::SCEoptim()` print convergence information to
> the console. The helper below captures that output and returns only the
> optimised value.

```{r benchmark-setup, eval = benchmark_packages_available}
#| code-fold: true
silent_rtop <- function(...) {
  utils::capture.output(
    res <- rtop::sceua(...),
    type = "output"
  )
  res$value
}

silent_soilhyp <- function(...) {
  utils::capture.output(
    res <- SoilHyP::SCEoptim(...),
    type = "output"
  )
  res$value
}

# The first `rtop` call in a fresh session can land in a different local basin;
# a single warm-up call makes the benchmarked calls comparable to `sceua`.
invisible(silent_rtop(
  OFUN,
  pars = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  maxn = 10000
))

invisible(silent_soilhyp(
  FUN = OFUN,
  par = c(0.1, 0.1, 0.1),
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  x = x,
  yobs = y,
  control = list(maxeval = 10000)
))
```

```{r benchmark, eval = benchmark_packages_available}
set.seed(1)

bench::mark(
  sceua = sceua::sceua(
    fn = OFUN,
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    initial = c(0.1, 0.1, 0.1),
    x = x,
    yobs = y,
    max_evaluations = 10000
  )$value,
  rtop = silent_rtop(
    OFUN,
    pars = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    maxn = 10000
  ),
  soilhyp = silent_soilhyp(
    FUN = OFUN,
    par = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    control = list(maxeval = 10000)
  ),
  iterations = 25,
  relative = FALSE,
  check = function(x, y) isTRUE(all.equal(x, y, tolerance = 1e-3))
)
```

## Q. Duan's test-function benchmarks

The following benchmarks use R translations of the Q. Duan test functions published alongside the [Matlab code](https://www.mathworks.com/matlabcentral/fileexchange/7671-shuffled-complex-evolution-sce-ua-method). We use one aligned strict configuration for
all problems: objective convergence is disabled and parameter convergence is
tightened to `1e-6`. Timing benchmarks are only run for packages that return
finite values and agree with at least one other package within `1e-3`. Spoiler, not all packages return finite values...

```{r duan-test-functions, eval = benchmark_packages_available}
#| code-fold: true
goldstein_price <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  u1 <- (x1 + x2 + 1)^2
  u2 <- 19 - 14 * x1 + 3 * x1^2 - 14 * x2 + 6 * x1 * x2 + 3 * x2^2
  u3 <- (2 * x1 - 3 * x2)^2
  u4 <- 18 - 32 * x1 + 12 * x1^2 + 48 * x2 - 36 * x1 * x2 +
    27 * x2^2
  (1 + u1 * u2) * (30 + u3 * u4)
}

rosenbrock <- function(x) {
  100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
}

six_hump_camelback <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  (4 - 2.1 * x1^2 + x1^4 / 3) * x1^2 + x1 * x2 +
    (-4 + 4 * x2^2) * x2^2
}

rastrigin_duan <- function(x) {
  x[1]^2 + x[2]^2 - cos(18 * x[1]) - cos(18 * x[2])
}

griewank_duan <- function(x) {
  divisor <- if (length(x) == 2L) 200 else 4000
  sum(x^2 / divisor) - prod(cos(x / sqrt(seq_along(x)))) + 1
}

shekel <- function(x) {
  a <- rbind(
    c(4, 1, 8, 6, 3, 2, 5, 8, 6, 7),
    c(4, 1, 8, 6, 7, 9, 5, 1, 2, 3.6),
    c(4, 1, 8, 6, 3, 2, 3, 8, 6, 7),
    c(4, 1, 8, 6, 7, 9, 3, 1, 2, 3.6)
  )
  c_values <- c(0.1, 0.2, 0.2, 0.4, 0.4, 0.6, 0.3, 0.7, 0.5, 0.5)

  -sum(vapply(seq_along(c_values), function(i) {
    1 / (sum((x - a[, i])^2) + c_values[i])
  }, numeric(1)))
}

hartman <- function(x) {
  a6 <- rbind(
    c(10, 0.05, 3, 17),
    c(3, 10, 3.5, 8),
    c(17, 17, 1.7, 0.05),
    c(3.5, 0.1, 10, 10),
    c(1.7, 8, 17, 0.1),
    c(8, 14, 8, 14)
  )
  c6 <- c(1, 1.2, 3, 3.2)
  p6 <- rbind(
    c(0.1312, 0.2329, 0.2348, 0.4047),
    c(0.1696, 0.4135, 0.1451, 0.8828),
    c(0.5569, 0.8307, 0.3522, 0.8732),
    c(0.0124, 0.3736, 0.2883, 0.5743),
    c(0.8283, 0.1004, 0.3047, 0.1091),
    c(0.5886, 0.9991, 0.6650, 0.0381)
  )

  -sum(vapply(seq_along(c6), function(i) {
    c6[i] * exp(-sum(a6[, i] * (x - p6[, i])^2))
  }, numeric(1)))
}

duan_control <- list(
  seed = 1L,
  complexes = 5L,
  max_evaluations = 10000L,
  kstop = 5L,
  pcento = 0,
  parameter_epsilon = 1e-6,
  tolerance = 1e-3
)

duan_problems <- list(
  goldstein_price = list(
    fn = goldstein_price,
    lower = c(-2, -2),
    upper = c(2, 2),
    initial = c(1, 1),
    optimum = 3
  ),
  rosenbrock = list(
    fn = rosenbrock,
    lower = c(-5, -5),
    upper = c(5, 5),
    initial = c(-3, 4),
    optimum = 0
  ),
  six_hump_camelback = list(
    fn = six_hump_camelback,
    lower = c(-5, -2),
    upper = c(5, 8),
    initial = c(0, 0),
    optimum = -1.031628453489877
  ),
  rastrigin_duan = list(
    fn = rastrigin_duan,
    lower = c(-1, -1),
    upper = c(1, 1),
    initial = c(0.5, 0.5),
    optimum = -2
  ),
  griewank_duan = list(
    fn = griewank_duan,
    lower = rep(-600, 10),
    upper = rep(600, 10),
    initial = rep(100, 10),
    optimum = 0
  ),
  shekel = list(
    fn = shekel,
    lower = rep(0, 4),
    upper = rep(10, 4),
    initial = rep(1, 4),
    optimum = -10.5364098252
  ),
  hartman = list(
    fn = hartman,
    lower = rep(0, 6),
    upper = rep(1, 6),
    initial = rep(0.5, 6),
    optimum = -3.322368011415515
  )
)

sceua_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  sceua::sceua(
    fn = problem$fn,
    lower = problem$lower,
    upper = problem$upper,
    initial = problem$initial,
    max_evaluations = duan_control$max_evaluations,
    kstop = duan_control$kstop,
    pcento = duan_control$pcento,
    complexes = duan_control$complexes,
    points_per_complex = 2L * n + 1L,
    simplex_size = n + 1L,
    evolution_steps = 2L * n + 1L,
    min_complexes = duan_control$complexes,
    parameter_epsilon = duan_control$parameter_epsilon
  )$value
}

rtop_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  silent_rtop(
    problem$fn,
    pars = problem$initial,
    lower = problem$lower,
    upper = problem$upper,
    maxn = duan_control$max_evaluations,
    kstop = duan_control$kstop,
    pcento = duan_control$pcento,
    ngs = duan_control$complexes,
    npg = 2L * n + 1L,
    nps = n + 1L,
    nspl = 2L * n + 1L,
    mings = duan_control$complexes,
    peps = duan_control$parameter_epsilon,
    iniflg = 1
  )
}

soilhyp_duan_value <- function(problem) {
  set.seed(duan_control$seed)
  n <- length(problem$lower)

  silent_soilhyp(
    FUN = problem$fn,
    par = problem$initial,
    lower = problem$lower,
    upper = problem$upper,
    control = list(
      maxeval = duan_control$max_evaluations,
      ncomplex = duan_control$complexes,
      cce.iter = 2L * n + 1L,
      reltol = duan_control$pcento,
      tolsteps = duan_control$kstop,
      initsample = "latin"
    )
  )
}

safe_duan_value <- function(package, problem) {
  value <- tryCatch(
    switch(
      package,
      sceua = sceua_duan_value(problem),
      rtop = rtop_duan_value(problem),
      soilhyp = soilhyp_duan_value(problem)
    ),
    error = function(e) NA_real_
  )

  as.numeric(value)
}

validate_duan_problem <- function(name, problem) {
  packages <- c("sceua", "rtop", "soilhyp")
  values <- vapply(packages, safe_duan_value, numeric(1), problem = problem)
  finite_values <- values[is.finite(values)]
  agrees <- length(finite_values) >= 2L &&
    isTRUE(all.equal(
      unname(finite_values),
      rep(unname(finite_values[1]), length(finite_values)),
      tolerance = duan_control$tolerance
    ))

  data.frame(
    #problem = name,
    #package = packages,
    value = values,
    error = values - problem$optimum,
    timed = is.finite(values) & agrees,
    stringsAsFactors = FALSE
  )
}

duan_validation <- do.call(
  rbind,
  Map(validate_duan_problem, names(duan_problems), duan_problems)
)

duan_timing_problems <- unique(duan_validation$problem[duan_validation$timed])

benchmark_duan_problem <- function(name, iterations = 3) {
  timed_packages <- duan_validation$package[
    duan_validation$problem == name & duan_validation$timed
  ]
  if (length(timed_packages) < 2L) {
    message("Skipping timing benchmark: fewer than two packages agree.")
    return(invisible(NULL))
  }

  problem <- duan_problems[[name]]
  expressions <- list(
    sceua = quote(sceua_duan_value(problem)),
    rtop = quote(rtop_duan_value(problem)),
    soilhyp = quote(soilhyp_duan_value(problem))
  )
  do.call(
    bench::mark,
    c(
      expressions[timed_packages],
      list(
        iterations = iterations,
        relative = FALSE,
        check = function(x, y) {
          isTRUE(all.equal(x, y, tolerance = duan_control$tolerance))
        }
      )
    )
  )
}
```

```{r duan-validation, eval = benchmark_packages_available}
duan_validation
```

### Goldstein-Price function

Bounds: `[-2, 2]` for both parameters. Global minimum: `f(0, -1) = 3`.

```{r benchmark-goldstein-price, eval = benchmark_packages_available}
benchmark_duan_problem("goldstein_price")
```

### Rosenbrock function

Bounds: `[-5, 5]` for both parameters. Global minimum: `f(1, 1) = 0`.

```{r benchmark-rosenbrock, eval = benchmark_packages_available}
benchmark_duan_problem("rosenbrock")
```

### Six-hump camelback function

Bounds: `[-5, 5]` for the first parameter and `[-2, 8]` for the second.
Global minimum: approximately `f(0.08983, -0.7126) = -1.031628`.

```{r benchmark-six-hump-camelback, eval = benchmark_packages_available}
benchmark_duan_problem("six_hump_camelback")
```

### Rastrigin-like function

Bounds: `[-1, 1]` for both parameters. Global minimum: `f(0, 0) = -2`.

```{r benchmark-rastrigin-duan, eval = benchmark_packages_available}
benchmark_duan_problem("rastrigin_duan")
```

### Griewank function

Bounds: `[-600, 600]` for all ten parameters. Global minimum:
`f(0, ..., 0) = 0`.

```{r benchmark-griewank-duan, eval = benchmark_packages_available}
benchmark_duan_problem("griewank_duan")
```

### Shekel function

Bounds: `[0, 10]` for all four parameters. Global minimum: approximately
`f(4, 4, 4, 4) = -10.5363`.

```{r benchmark-shekel, eval = benchmark_packages_available}
benchmark_duan_problem("shekel")
```

### Hartman function

Bounds: `[0, 1]` for all six parameters. Global minimum: approximately
`f(0.201, 0.150, 0.477, 0.275, 0.311, 0.657) = -3.3223`.

```{r benchmark-hartman, eval = benchmark_packages_available}
benchmark_duan_problem("hartman")
```
