---
title: "Benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sceua)
```

## A non-linear curve-fitting problem

This problem is adapted from the `rtop::sceua()` documentation. We generate
noisy observations from a sine function with three parameters and recover them
by minimising the sum of squared errors.

```{r sine-data}
set.seed(1)

fun <- function(x, pars) pars[2] * sin(x * pars[1]) + pars[3]

x <- rnorm(50, sd = 3)
y <- fun(x, pars = c(5, 2, 3)) + rnorm(length(x), sd = 0.3)

OFUN <- function(pars, x, yobs) {
  yvals <- fun(x, pars)
  sum((yvals - yobs)^2)
}
```

## Timing comparison

The `rtop` package contains a pure-R SCE-UA implementation. We can compare it
with the Rust-backed `sceua` package on the same problem using the `bench`
package.

`rtop::sceua()` prints convergence information to the console. The helper below
captures that output and returns only the optimised value.

```{r benchmark, eval = requireNamespace("rtop", quietly = TRUE) && requireNamespace("bench", quietly = TRUE)}
# By default, rtop::sceua() prints convergence information to the console which clatters the output.
silent_rtop <- function(...) {
  utils::capture.output(
    res <- rtop::sceua(...),
    type = "output"
  )
  res$value
}

bench::mark(
  sceua = sceua::sceua(
    fn = OFUN,
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    initial = c(0.1, 0.1, 0.1),
    x = x,
    yobs = y,
    max_evaluations = 10000,
    seed = 1
  )$value,
  rtop = silent_rtop(
    OFUN,
    pars = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y,
    maxn = 10000
  ),
  iterations = 5,
  check = TRUE
)
```
