---
title: "Getting started with sceua"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with sceua}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`sceua` provides R bindings to a Rust implementation of the Shuffled Complex
Evolution - University of Arizona (SCE-UA) global optimisation algorithm
(Duan et al., 1992). It is designed to be lightweight and self-contained:
the only runtime dependency is `checkmate`.

## A first example

The objective function must accept a single numeric vector and return a
scalar. Here we minimise the two-dimensional sphere.

```{r sphere}
library(sceua)

result <- sceua(
  fn = function(x) sum(x^2),
  lower = c(-5, -5),
  upper = c(5, 5),
  max_evaluations = 5000,
  kstop = 5,
  pcento = 1e-8,
  seed = 1969,
  complexes = 5
)

result
```

## Fitting a sine curve

This example is adapted from the `rtop::sceua()` documentation. We generate
noisy data from a sine function with three parameters and recover them by
minimising the sum of squared errors.

```{r sine-data}
set.seed(1)

fun <- function(x, pars) pars[2] * sin(x * pars[1]) + pars[3]

x <- rnorm(50, sd = 3)
y <- fun(x, pars = c(5, 2, 3)) + rnorm(length(x), sd = 0.3)

OFUN <- function(pars, x, yobs) {
  yvals <- fun(x, pars)
  sum((yvals - yobs)^2)
}

result <- sceua(
  fn = OFUN,
  lower = c(-10, 0, -10),
  upper = c(10, 10, 10),
  initial = c(0.1, 0.1, 0.1),
  x = x,
  yobs = y,
  max_evaluations = 10000,
  seed = 1969
)

result
```

## Benchmark against `rtop::sceua()`

The `rtop` package contains a pure-R SCE-UA implementation. If `rtop` is
installed, we can compare the two on the sine-curve problem.

```{r benchmark, eval = requireNamespace("rtop", quietly = TRUE)}
rtop_time <- system.time({
  rtop_result <- rtop::sceua(
    OFUN,
    pars = c(0.1, 0.1, 0.1),
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    x = x,
    yobs = y
  )
})

sceua_time <- system.time({
  our_result <- sceua(
    fn = OFUN,
    lower = c(-10, 0, -10),
    upper = c(10, 10, 10),
    initial = c(0.1, 0.1, 0.1),
    x = x,
    yobs = y,
    max_evaluations = 10000,
    seed = 1969
  )
})

data.frame(
  package = c("rtop", "sceua"),
  elapsed = c(rtop_time["elapsed"], sceua_time["elapsed"]),
  value = c(rtop_result$value, our_result$value)
)
```

## References

Duan, Q., Sorooshian, S., and Gupta, V.K., 1992. Effective and efficient
global optimization for conceptual rainfall-runoff models.
*Water Resources Research* 28 (4), 1015-1031.
